---
title: "Day 3 Analyzing Experiments in R Template"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

This template has two sections. In the first section you will work with experimental data `ncp_vax.RData`. You need `Day 3.html` to guide you through the questions.

In the second section, you will work with `STAR.csv`, which is also an experimental data. This section is derived from Chapter 2, "Estimating Causal Effects with Randomized Experiments" (p. 27 - 50).

Before doing anything, install and load the packages required for plotting.

```{r setup, include=FALSE}
# this block will not show up in the knitted file because of argument 'include=FALSE'

# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org/"))

install.packages('ggplot2')
install.packages('gplots')

library(ggplot2)
library(gplots)
```

### Do Different Vacinne Appeals Motivate Vaccine Uptake?

1.  Load `ncp_vax.RData`. What is the unit of analysis? What is the dimension of the data? 

```{r}
load('ncp_vax.RData')
```

2. Read `Day 3.html` carefully. What do the variables in the dataset mean? What are the outcome variable and the treatment variables in this dataset?


3. How many respondents were assigned to the treatment and control groups?

```{r}
# code here
```

4. For experiments to work, the treatments should be randomly assigned. How would you check whether the treatment assignment is random? (Hint: Check whether distribution of gender and education is similar across treatment and control groups)

```{r}
# code here
```

5. Find the average treatment effect (ATE) by computing the difference in means estimator.

```{r}
# code here
```

6. You would want to test whether the effect of the treatment is statistically significant. Compute the difference in means and use t-test to check statistical significance. What is the null hypotheses and the alternative hypotheses? Add interpretation of the results.

```{r}
# code here
```

7. Visualize the group means.

```{r}
# code here
```

8. Visualize the average treatment effects and the confidence intervals.

```{r}
# code here
```

### Do Small Classes Improve Student Performance?

In this section, you will work with `STAR.csv` data, which is a portion of the data collected from the experiment called ProjectSTAR. The aim of the project was to examine the effects of class size on student performance in both the short and long term. The project consisted of an experiment conducted in 1980 in which kindergartners in the state of Tennesssee, USA, were randomly assigned to attend either small classes, with 13 to 17 students, or regular-size classes, with 22 to 25 students, until the end of third grade.  Researchers followed student progress overtime. As the outcome variables of interest, we have student scores on third-grade standardized tests in reading and math as well as high school graduation rates.

Each variable in the data set mean the following:

- `classtype` : whether the class is regular or small
- `reading`: students' reading scores
- `math`: students' math scores
- `graduated`: whether the students ended up graduating highschool (`1=graduated`, `0=did not graduate`)

1. Read the STAR data

```{r}
star<-read.csv('STAR.csv')
```

2. What is the dimension of the data? What is the data type of each variable?

```{r}
# code here
```

3. What are the outcome variables and the treatment variable in the experiment?

```{r}
# code here
```

4. Find the average treatment effect (ATE) by computing the difference in means estimator. 

```{r}
# code here
```

Do you need to recode any variables? If so, recode the variables.

```{r}
# code here
```

5. You would want to test whether the effect of the treatment is statistically significant. Compute the difference in means and use t-test to check statistical significance. What is the null hypotheses and the alternative hypotheses? Add interpretation of the results.

```{r}
# code here
```

6. Visualize the group means.

```{r}
# code here
```

7. Visualize the average treatment effects and the confidence intervals.

```{r}
# code here
```

