---
title: "Day 2: Regression in R Template"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

We will continue to work with `ncp_min.RData`. You will also need the `Day-2.html` as a guide to work on this template. We will work through the example of whether left-right self-placement (`lr_scale`) predicts immigration attitudes (`immigration`). 


Before starting with anything, download the packages and load using `library()`:
```{r}
install.packages("stargazer")
install.packages("sjPlot")

# loading the packages
library(stargazer)
library(sjPlot)
```

Make sure to have everything in the same folder - the .RData, and this .Rmd file. Or else you will not be able to knit.
 
1. Load `ncp_min.RData`

```{r}
load('ncp_min.RData')
```

2. Are there any missing values in the data? If so, remove the missing values and assign a new name to the dataframe after removing the rows containing the missing values.

```{r}
# write your code here
```

3. Let's say you want to test whether right-wing self-placement is associated with more restrictive immigration attitudes. What is your null hypothesis and the alternative hypothesis? What is your outcome variable and the predictor variable?


4. One way to asses the relationship between the two variables you are interested in is visualizing it. Visualize the relationship between right-wing self-placement and immigration attitudes. Which plot will you use? (scatter plot, bar plot, boxplot...)

```{r}
# write your code here
```

5. You can also asess the relation by calculating the correlation using `cor()`. What is the correlation coefficient? What is the strength and the direction of the relationship between `lr_scale` and `immigration`?

```{r}
# write your code here
```

6. Fit a univariate linear regression model using `lm()`, use `summary()` to check the results. Interpret the results.

```{r}
# write your code here
```


What is the intercept and the coefficient estimate of `lr_scale`? Explain how one unit increase in the predictor variable `lr_scale` changes the outcome variable `immigration`, and the statistical significance of the results. Can you reject the null hypothesis or fail to reject the null hypothesis? On what grounds? 

```{r}
# write your code here
```



7. What is the predicted value of immigration for respondents who place themselves as a 0 (far left) on the `lr_scale`? How about those who place themselves `lr_scale = 5` (center) and `lr_scale = 10` (far right)?

```{r}
# write your code here
```

8. Asses the fit of the model. (R-squared, adjusted R-squared, visualization..)

One way to assess the fit of a liner regression model is plotting the linear regression model result on a scatter plot. First plot your predictor and outcome variable and add the regression line using `abline()`. Use `jittter()` for better visualization.

9. Fit a multivariate model, which also includes other variables such as year of birth, gender, and education. These variables are called 'covariates' or 'control variables' as they could also affect the variation in the outcome variable `immigration`. Use `summary()` to see the results.

```{r}
# write your code here
```

10. Print the univariate and multivariate linear regression results in a table using `stargazer()`

```{r}
# write your code here
```

11. According to the multivariate model, which variables explain the variation in the outcome variable `immigration()` and how? Make sure to note the magnitude of the coefficient, the direction (positive or negative), and the statistical significance.

```{r}
# write your code here
```

12. How does the coefficient estimate `lr_scale` in the univariate model compare to that in the multivariate model? If there is a difference, why?


13. How does the Adjusted R² of the multivariate model compare to the univariate model? Does adding predictors improve the model’s fit?


14. Visualize coefficient estimates and the confidence intervals for the univariate and the multivariate model. Use `plot_models()`

```{r}
# write your code here
```

Knit the .Rmd file to .html file and send it to either Soran or Esther for feedback.

Also if you have time left: come up with a research question that can be answered with the data we have. The only rule is that you are not allowed to use immigration as the outcome variable. Formulate a hypothesis and test it using methods we have covered in this session.

Conduct the analyses in a separate rmarkdown document, where you explain (very succinctly) what your hypothesis is, and the theory behind it. Tabulate and/or plot your results. Email one of us the output (tables and/or figures) so that we can go through them together at the end (if we have time).