---
title: "Assignments Day 1 (answer key)"
author: "Soran Hajo Dahl"
date: "`r Sys.Date()`"
output: html_document
---

### Part 1: 5+3

What is 5+3?

```{r}
5+3
```

### Part 2: Load up and examine the data

For the rest of the questions, we'll we'll use the same data as in the rest of the session. Load it up in the code chunk below, and find the following:

-   What are the variables called, and what class (e.g. `character`) is each of them?
```{r}
load("ncp_min.RData")
str(ncp_min)
```

Gender = chr, education = num, yob = num, vote_intention = chr, lr_scale = num, immigration = num, eu = num, climate = num, economy = num

-   Look up the first five observations. What is the gender of the third individual from the top?

```{r}
head(ncp_min)
```

The third responent from the top is female.

### Part 3: Tables
-   How many of the respondents have higher education?

```{r}
table(ncp_min$education)

#Alternatively:
sum(ncp_min$education == 3, na.rm = TRUE)
```
1299 respondents have higher education. 

-   Recode the education variable so that those without a university degree (`education` = 1 or 2) are combined into a single category coded 0, while respondents with a university degree (`education` = 3) are codeded 1. Then find the *percentage* of respondents that do not have a university degree.

```{r}
ncp_min$education <- ifelse(ncp_min$education == 3, 1, 0)
prop.table(table(ncp_min$education))
```
0.35% of respondents do not have a university degree.

-   Which party has the highest share of university degree holding voters? Which party has the lowest? What percentage of these parties' voters have a university degree?

```{r}
prop.table(table(ncp_min$vote_intention, ncp_min$education), 1)
```

MDG has the highest share of degree-holding voters (89%) while Frp has the lowest share (42%)

### Part 4: Averages
-   On average, are the respondents mostly positive or negative towards EU membership?

```{r}
mean(ncp_min$eu, na.rm = TRUE)
```
The respondents are slighty more negative than positive towards EU membership.

-   On average, are degree holders or non-degree holders more positive towards EU membership? What, exactly, are their positions on average?
```{r}
# Mean attitude towards EU for degree holders
mean(ncp_min$eu[ncp_min$education == 1], na.rm = TRUE)

# Mean attitude towards EU for non-degree holders
mean(ncp_min$eu[ncp_min$education == 0], na.rm = TRUE)
```
Degree holders are more positive (3.8) towards EU membership than non-degree holders (4.6), on average. 


### Part 5: Visualizing data
-   Create a barplot of the respondents attitudes towards EU-membership:

```{r}
barplot(table(ncp_min$eu),
        ylab = "Counts", xlab = "Party", main = "Vote Intention Counts")

```

-   Create a barplot of the respondents attitudes towards EU-membership broken down by their level of education (for extra points, plot the percentages rather than frequencies):

```{r}
barplot(prop.table(table(ncp_min$education, ncp_min$eu), 1),
        beside = TRUE, legend.text = TRUE,
        ylab = "Percentages", xlab = "Attitude towards EU-membership (1 = positive, 7 = negative)", main = "Attitudes towards EU-membership by level of education")


```

-   Now create boxplot showing attitudes towards EU-membership by vote intention.

```{r}
boxplot(ncp_min$eu~ncp_min$vote_intention)
```

### Part 6: Correlation

-   What is the correlation between attitudes towards EU membership and immigration attitudes? Write a short interpretation of the result.

```{r}
cor(ncp_min$eu, ncp_min$immigration, use = "complete.obs")
```
The correlation is moderately positive (0.21). This means that higher values of `eu` (indicating opposition to EU membership) are associated with higher values of `immigration` (indicating negative views of immigration). Accordingly, anti-EU sentiments are positively associated with anti-immigration sentiments.