---
title: "Day-0 Practice Questions"
output: html_document
date: "2025-01-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Try answering the following questions using the British Election Study data.

1. Which variables contain missing values? 
2. How many 'don't know' responses are in the dataset?
3. What is the percentage of respondents who answered 'leave'?
4. What is the percentage of respondents who answered 'don't know'? 
5. Plot the distribution of answers in the 'vote' column?
6. What is the average age of the respondents?
7. What percent of respondents in the data set are over 60? (tip: Use logical operators `>, <, ==, !=, <=, >=`)
8. What percent of respondents in the data set are exactly 60 years old?
9. What percent of respondents in the data set are NOT 60 years old?
10. What is the distribution of age? Do you notice anything? (tip: use `hist()` and `boxplot()`)
11. Examine distribution of age of brexit supporters and non-supporters. What do you notice?
12. Check distribution of brexit non-supporters by education levels.

```{r}
bes<-read.csv('BES.csv')

# examine the data --
head(bes)
str(bes)
```
